---
title: "Tidyverse Wrangling with Factors"
subtitle: "factors, levels, categorical data w/ base & `forcats`"
date: "2/10/2026"
footer: "D2M-R I | Week 6 & 7" #  quarter, week, and class

execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
set.seed(123)
```

# Introduction to Factors

## Categorical Data


Categorical variables

-   Represent a set of distinct groups or categories that are:
-   Finite in number
-   Mutually exclusive
-   Non-numeric
    -   i.e., mathematical operations don't make sense

Examples

-   Gender
-   Education level
-   Survey responses (e.g., "Agree", "Neutral", "Disagree")


## Categorical Data Types


Category Types

-   **Nominal**: No inherent order (e.g., colors, types of fruit)
-   **Ordinal**: Meaningful order (e.g., "Low", "Medium", "High")

Also known as

-   Factor variables
-   Discrete variables
-   Qualitative data 
    -   Terrible. Don't call it that.


## Factor data

-   [Factors]{.keyword} are R's representation of categorical data
-   Integer vectors with level labels

```{r}
#| label: factor-structure

f <- factor(c("low", "medium", "high", "medium", "low"))

typeof(f)
str(f)
```



## Factors vs. strings

Factors

- Stored as integers with labels
- Memory efficient (at large scale)
- Control ordering
- Useful for: 
    -   non-numeric modeling
    -   visualization

Strings

- Stored as many ordered character units
- More memory (at large scale)
- Flexible manipulation
- Useful for: 
    -   text processing
    -   unique values

## Memory efficiency

Compare object size of factor vs character vector with repeated values

```{r}
#| label: factor-vs-string-small-define

# Small scale 
small_strings <- c(
    "low", "medium", 
    "high", "medium", "low")
small_factor <- factor(small_strings)

```

```{r}
#| label: factor-vs-string-small-compare

# Strings: ~.3 KB
object.size(small_strings) 

# Factor: ~.7 KB - negligibly larger
object.size(small_factor) 
```

```{r}
#| label: factor-vs-string-large-define

# Larger scale
big_strings <- rep(c(
    "low", "medium", "high"), 
    length.out = 10000) # repeat to length 10,000
big_factors <- factor(big_strings)

```

```{r}
#| label: factor-vs-string-larger-compare

# Strings: ~80.2 KB
object.size(big_strings)  

# Factor:  ~40.6 KB - ~50% smaller
object.size(big_factors)  
```

```{r}
#| label: factor-vs-string-long-strings-define

# Long strings
long_strings <- rep(c(
    "Dissatisfied, very dissatisfied, or extremely dissatisfied", 
    "Not at all dissatisfied, somewhat dissatisfied, or a little dissatisfied", 
    "Neutral and/or no response given"), 
    length.out = 2000) # repeat to length 2,000
long_factors <- factor(long_strings)
```

```{r}
#| label: factor-vs-string-long-strings-compare

# Strings: ~8.4 KB
object.size(long_strings) 

# Factor: ~4.8 KB - ~50% smaller
object.size(long_factors)  
```


## Base R factor functions

**Core base R factor functions:**

- `factor()`: Create factor
- `levels()`: View/set levels
- `nlevels()`: Count levels
- `as.character()`: Convert to string
- `as.numeric()`: Get underlying integers


```{r}
#| label: base-r-factors

f <- factor(c("low", "medium", "high", "medium", "low"))

levels(f)  # View levels
nlevels(f) # Count levels
as.character(f)  # Convert to strings
as.numeric(f)  # Get underlying integers
```


## `forcats` package: simplify working with factors {.left-sidebar-slide fullscreen=true style="font-size: .8;"}


[forcats]{.pkg}: <br> Simplify working with factors

-   Tidyverse package _**for cat**_egorical variables
-   Largely helpers/wrapper for base R functions with:
    -   Consistent `fct_*` naming
    -   More intuitive syntax
    -   Better compatibility with tidyverse workflows
    -   Additional functionality for common tasks

Create: `as_factor`
Inspect: `fct_unique`, `fct_count`, `fct_match`
Combine: `fct_c`, `fct_unify`
Reorder: `fct_relevel`, `fct_infreq`, `fct_reorder`, `fct_reorder2`, `fct_rev`, `fct_inorder`, `fct_shift`, `fct_shuffle`
Reassign: `fct_recode`, `fct_relabel`, `fct_anon`, `fct_collapse`, `fct_lump_*n`, `fct_other`
Add/Drop: `fct_expand`, `fct_drop`, `fct_na_value_to_level`


# Create Factors

## Create factors

-   **Base:** Create factors with `factor()`
    -   Alphabetizes levels by default
    -   Includes arguments to specify order, labels, and more

```{r}
#| label: create-factors-factor

# Using factor()
f1 <- factor(c("low", "medium", "high", "medium", "low"))
levels(f1)  # Alphabetical order
as.numeric(f1)  # high=1, low=2, medium=3
```

-   **`forcats`:** Create factors with `as_factor()`
    -   Preserves order of appearance
    -   No additional specification needed or possible


```{r}
#| label: create-factors-as-factor

# Using as_factor() from forcats
f2 <- as_factor(c("low", "medium", "high", "medium", "low"))
levels(f2)  # Order of appearance
as.numeric(f2)  # low=1, medium=2, high=3
```


## Create: specify level order


-   Define levels when creating factors: `f <- factor(x, levels = c(...))`
    -   `levels` argument: Set specific levels and order
    -   Overrides default (alphabetical)
    -   Elements of `levels` vector must match unique values in `x`
    -   Levels not in `x` will be created but have no observations
    -   Observations in `x` without corresponding levels will become NA
-   *no forcats function*
    -   `as_factor()` necessarily preserves order of appearance


```{r}
#| label: specify-levels-factor

x <- c("B", "A", "C", "A", "B")
f <- factor(x, levels = c("C", "B", "A"))
f
levels(f)  # Custom order
```


## Create: specify level labels

`factor(x, levels = c(...), labels = c(...))`

```{r}
#| label: specify-labels-define

x <- c("B", "A", "C", "A", "B")
```


```{r}
#| label: specify-levels-only

f_levels_only <- factor(x, 
    levels = c("C", "B", "A"))

f_levels_only # Custom order, default labels
levels(f_levels_only)  

```

```{r}
#| label: specify-with-labels

f_with_labels <- factor(x, 
    levels = c("C", "B", "A"), 
    labels = c("Gamma", "Beta", "Alpha"))
f_with_labels # Custom order & labels

```


## Create: specify meaningful order


**For ordinal data with meaningful order**

- Likert scales (Strongly Disagree → Strongly Agree)
- Education levels (HS → Bachelor's → Master's → PhD)
- Sizes (Small → Medium → Large)


## Create: ordered vs. unordered factors

`factor(..., ordered = TRUE)`: 

Encode order with argument

```{r}
#| label: ordered-factors-argument

sizes_unordered <- factor(
    c("medium", "small", "large", "small", "medium"),
    levels = c("small", "medium", "large"),
    ordered = TRUE)
sizes_unordered
levels(sizes_unordered)
as.numeric(sizes_unordered) 
```

`ordered(...)`: 

Create ordered factor directly

```{r}
#| label: ordered-factors-direct

sizes_ordered <- ordered(
    c("medium", "small", "large", "small", "medium"),
    levels = c("small", "medium", "large"))
sizes_ordered
levels(sizes_ordered)
as.numeric(sizes_ordered) 

```

:::

::::

::: notes

Levels always have an order, because they are stored as integers and integers are necessarily ordered.
*Ordered* factors explicitly encode that order for ordinal data. This allows for meaningful comparisons and appropriate statistical treatment in models.

When you actually view the contents of an ordered factor, you'll see that its levels aren't just listed in order, they have `<` between them. 
This is saying to the human user (and R functions that respect ordered factors) that there is a meaningful order to these levels, not simply an arbitrary assignment to numerals that happen to have an order because that's how numbers work.

::: 


## Meaningful ordering


```{r}
#| label: unordered-factor-order

sizes_unordered <- factor(
    c("medium", "small", "large", "small", "medium"),
    levels = c("small", "medium", "large"),
    ordered = FALSE)

sizes_unordered

```

```{r}
#| label: unordered-factor-comparisons
#| warning: false

sizes_unordered[1] < sizes_unordered[3]  
sizes_unordered[2] > sizes_unordered[4]  
sizes_unordered[2] == sizes_unordered[4] 
```


```{r}
#| label: ordered-factor-order

sizes_ordered <- factor(
    c("medium", "small", "large", "small", "medium"),
    levels = c("small", "medium", "large"),
    ordered = TRUE) # or use ordered() function

sizes_ordered


```


```{r}
#| label: ordered-factor-comparisons
#| warning: false

sizes_ordered[1] < sizes_ordered[3]  
sizes_ordered[2] > sizes_ordered[4]  
sizes_ordered[2] == sizes_ordered[4] 
```

## Inspect factors


**Examine existing factor structure**


```{r}
#| label: create-factor-inspect
x <- factor(c("medium", "high", "low", "high", "high", "medium"))
```

- Base:
    - `levels()`: View level labels
    - `nlevels()`: Count levels

```{r}
#| label: inspect-levels-base

levels(x) # character vector
nlevels(x) # integer count
```

- `forcats`:
    - `fct_unique()`: Unique values of levels
    - `fct_count()`: Frequency table  
    
```{r}
#| label: inspect-levels-forcats
fct_unique(x) # factor of levels
fct_count(x) # tibble of lvl & ct.
```


## Combine factors

```{r}
#| label: combine-factors-define

f1 <- factor(c("a", "c"))
f2 <- factor(c("b", "a"))
```


`fct_c()`: Combine factors

```{r}
#| label: combine-factors-fct-c

f_combined <- fct_c(f1, f2)

f1
f2
f_combined

```


`fct_unify()`: Unify levels of factors

```{r}
#| label: combine-factors-fct-unify

unified_list <- list(f1, f2)
f_unified <- fct_unify(unified_list)

f_unified

```


# Reorder Factor Levels

## Reorder: manual assignment (relevel)

`fct_relevel()`: Move levels to specific positions

```{r}
#| label: manual-reordering-create
#| output-location: column

f <- factor(c("medium", "low", "high", "medium", "low"))
f # Current order
```


Move specific level to front

```{r}
#| label: manual-reordering-move-front

# Move "low" to front
fct_relevel(f, "low")  
```


Move one level relative to another

```{r}
#| label: manual-reordering-relative-position

# Move "medium" before "high"
fct_relevel(f, "medium", "high")  
```


Move one level to specific position

```{r}
#| label: manual-reordering-absolute-position

# Move "high" after first level
fct_relevel(f, "high", after = 1)  

# Move "low" to last position
fct_relevel(f, "low", after = Inf)
```


## Reorder: reverse

`fct_rev()`: Reverse current level order

```{r}
#| label: reverse-order-create

f <- factor(c("large", "small", "medium", "medium", 
    "small", "medium", "small", "medium", "medium", "medium"), 
    levels = c("small", "medium", "large"))
f
```

```{r}
#| label: reverse-order-rev
fct_rev(f)
```

## `fct_rev` use case: flip bar chart

**Use case**: Flip bar chart from top-to-bottom to bottom-to-top


```{r}
ggplot(data.frame(f), aes(x = f)) + 
  geom_bar() +
  labs(title = "Bar Chart with Original Factor Order")
```


```{r}
ggplot(data.frame(f), aes(x = fct_rev(f))) +
  geom_bar() +
  labs(title = "Bar Chart with Reversed Factor Order")
```


## Reorder: by appearance

`fct_inorder()`: Order by first appearance

```{r}
#| label: reorder-by-appearance-create
#| output-location: column

f <- factor(c("b", "b", "a", "c", "c", "c"))
f
```


Reorder by first appearance

```{r}
#| label: reorder-by-appearance-inorder

fct_inorder(f)
```


Equivalent to creating with `as_factor()`

```{r}
f <- as_factor(c("b", "b", "a", "c", "c", "c"))
f
```


## Reorder: by frequency

`fct_infreq()`: Most common first

```{r}
#| label: reorder-by-frequency-create
f <- factor(c("b", "b", "a", "c", "c", "c"))
f
```

Reorder by frequency, most common first

```{r}
fct_infreq(f)
```


## `fct_infreq` use case: interpretable bar charts

```{r}
#| label: bar-chart-factor-create

f <- factor(c("b", "b", "b", "a", "c", "c"))
```


```{r}
#| label: bar-chart-default-ordered

ggplot(data.frame(f), aes(x = f)) +
  geom_bar() +
  labs(title = 
          "Bar Chart with Default (Alphabetical) Ordered Factor")
```


```{r}
#| label: bar-chart-infreq-ordered

ggplot(data.frame(f), aes(x = fct_infreq(f))) +
  geom_bar() +
  labs(title = 
          "Bar Chart with Frequency-Ordered Factor")
```


## Reorder: by another variable (summary)

```{r}
#| label: create-small-diamonds
#| include: false

diamonds_100 <- diamonds |> 
    filter(carat < 1.2) |> 
    slice_sample(n = 100)
```


`fct_reorder(f, x, fun)`: Order factor by summary of another variable 

Default: sort by level median

```{r}
#| label: reorder-by-other-variable-median

ggplot(diamonds_100, aes(
    x = fct_reorder(cut, carat), y = carat)) +
    geom_boxplot() +
    labs(title = "Reordered by Median Carat Value")
```


Or sort by any other summary function, e.g., `max`

```{r}
#| label: reorder-by-other-variable-max

ggplot(diamonds_100, aes(
    x = fct_reorder(cut, carat, max), y = carat)) +
    geom_boxplot() +
    labs(title = "Reordered by Maximum Carat Value")
```


## Reorder: by another variable (relationship)

`fct_reorder2(f, x, y)`: Order by relationship with two variables

Compare last values of y for each level of f across x (great for line plots)

```{r}
#| label: reorder-by-other-variable-reorder2
#| output-location: column

diamonds_100 |> 
    ggplot(aes(x = carat, y = price, 
        color = fct_reorder2(cut, carat, price))) +
    geom_line(stat = "summary", fun = mean) +
    labs(title = "Line Plot with Legend Ordered by Endpoint")
```


# Reassign Factor Levels

## Reassign: rename levels

`fct_recode()`: Manually rename specific levels `new_name = "old_name"`

Create factor with messy level names

```{r}
#| label: rename-levels-create

f <- factor(c("Low ", "Medium", "High", "Medium ", "Low", "high", "Hihg"))
f
```


Clean up names. Map multiple old levels to same new or existing level individually.

```{r}
f_clean <- fct_recode(f,
                      Low = "Low ",
                      Medium = "Medium ",
                      High = "high",
                      High = "Hihg")

f_clean
```


## Reassign: collapse levels manually

`fct_collapse()`: Combine multiple levels into one `new_name = c("old1", "old2")`

```{r}
#| label: collapse-levels-create

f <- factor(c("A", "A", "B", "B", "B", "C", "C", "C", "C", "D", "E", "F", "G", "G", "G", "G", "G"))

```


```{r}
#| label: collapse-levels-manual
#| output-location: column

f_collapsed <- fct_collapse(
    f,
    Group1 = c("A", "B"),
    Group2 = c("C", "D"),
    Group3 = "E")

f_collapsed
```

## Reassign: lump levels manually

`fct_other()`: Lump (non/)specified levels into "Other"

```{r}
#| label: lump-levels-create

f <- factor(c("A", "A", "B", "B", "B", "C", "C", "C", "C", "D", "E", "F", "G", "G", "G", "G", "G"))
```

```{r}
#| label: lump-levels-manual-keep

fct_other(f, keep = c("A", "B", "C"))
```

```{r}
#| label: lump-levels-manual-drop

fct_other(f, drop = c("A", "B", "C"))
```


## Reassign: lump levels automatically

`fct_lump_*()` family: Keep levels by condition, lump rest into "Other"

```{r}
#| label: collapse-levels-create-dup

f <- factor(c("A", "A", "B", "B", "B", "C", "C", "C", "C", "D", "E", "F", "G", "G", "G", "G", "G"))

```

`fct_lump_n()`: keep most frequent N

```{r}
#| label: collapse-levels-lump-n

fct_lump_n(f, n = 3)
```


`fct_lump_prop()`: keep levels above percent frequency

```{r}
#| label: collapse-levels-lump-prop

fct_lump_prop(f, prop = 0.2)
```

`fct_lump_min()`: keep levels with at least N observations

```{r}
#| label: collapse-levels-lump-min

fct_lump_min(f, min = 4)
```



# Add/Drop Levels

## Add/Drop: add and remove levels


```{r}
#| label: add-levels-create

f_abc <- factor(c("A", "B", "C"))
f_abc
```

`fct_expand()`: Add new levels without observations

```{r}
#| label: add-levels-expand

fct_expand(f, "D", "E")
```

`fct_drop()`: Remove levels with no observations

```{r}
#| label: drop-levels-create

f_abcde <- factor(c("A", "B", "C"), 
    levels = c("A", "B", "C", "D", "E"))
f_abcde

```

```{r}
#| label: drop-levels-drop

fct_drop(f_abcde)
```

In combination

```{r}
#| label: remove-again

fct_expand(f_abc, "D", "E") |> fct_drop()
```


## `fct_drop` use case: tibble filter

```{r}
#| label: drop-create-tibble
#| output-location: column

# Create a tibble with a factor variable
t <- tibble(
    f = factor(c("A", "B", "C", "D", "E")),
    i = c(1:5)
)

t
```

```{r}
#| label: drop-view-factor-column
#| output-location: column

t$f
```


```{r}
#| label: drop-unused-levels-filter
#| output-location: column


t_subset <- filter(t, i <= 3)
t_subset$f

```

```{r}
#| label: drop-unused-levels-droplevels
#| output-location: column

fct_drop(t_subset$f)
```



## Add/Drop: handle missing values

`fct_na_value_to_level()`: Convert between `NA` values and `NA` levels

```{r}
#| label: handle-missing-values-create

f <- factor(c("A", "B", NA, "A", "C", NA, "B"))
f
```


Create explicit NA level

```{r}
#| label: handle-missing-values-na-level

fct_na_value_to_level(f)
```

Convert NAs to named level

```{r}
#| label: handle-missing-values-named-level

fct_na_value_to_level(f, level = "(Missing)")
```



## See Also 

`forcats` functions we didn't talk about:

::: {.nonincremental}
1.  Create: n/a -- just `as_factor()`
2.  Inspect:
    1.  `fct_match(f, lvls)`: check for levels in f 
3.  Combine:
    1.  `fct_cross(f1, f2)`: Create interaction factor
4.  Reorder
    1.  `fct_inseq()`: Order by natural sequence
    2.  `fct_shift()`: Shift levels L or R, wrapping around
    3.  `fct_shuffle()`: Randomly permute levels
5.  Modify
    1.  `fct_lump_lowfreq()`: Lump low-frequency levels
    2.  `fct_relabel()`: Rename levels with function
    3.  `fct_anon()`: Anonymize to random integers

:::


# Practical Applications

## Data cleaning workflow

**String cleaning → Factor creation**

Common workflow:

1. Trim whitespace
2. Standardize case
3. Fix inconsistencies
4. Convert to factor

Messy categorical (string) data with inconsistent entries

```{r}
#| label: messy-categorical-data

# Create messy data
messy <- c(" Low", "medium ", "HIGH", "medium", " low", "high", NA, "Hihg")
```


Pipeline wth `str` and factor creation in one step

```{r}
#| label: cleaning-pipeline

messy |> 
    str_trim() |> 
    str_to_lower() |> 
    str_replace("hihg", "high") |> 
    factor(levels = c("low", "medium", "high"), 
        ordered = TRUE) |> 
    fct_na_value_to_level("no response")
```


## Visualization preparation

**Make better plots with factor control**

- Reorder for sorted bars (`fct_infreq`, `fct_reorder`)
- Collapse rare categories (`fct_lump_*`)
- Control legend order

## Visualization preparation

Create a simple bar chart

```{r}
#| label: bar-chart-unsorted
#| echo: true
#| eval: false

diamonds_100 |> 
    ggplot(aes(x = clarity)) +
    geom_bar() +
    labs(title = "Bar chart with no ordering",
        x = "Default factor variable")
```

Reorder factor by decreasing frequency

```{r}
#| label: bar-chart-freq
#| echo: true
#| eval: false

diamonds_100 |> 
ggplot(aes(x = 
        fct_infreq(clarity))) +
geom_bar() +
labs(title = "Bar chart with freq. ordering",
    x = "Somewhat arranged factor variable")
```

Lump rare categories for cleaner visualization

```{r}
#| label: bar-chart-freq-lump
#| echo: true
#| eval: false

diamonds_100 |> 
ggplot(aes(x = 
        fct_lump_n(clarity, n = 4) |> 
        fct_infreq())) +
geom_bar() +
labs(title = "Bar chart with freq. 
    ordering + lumping",
    x = "Cleverly arranged factor variable")
```


Move "other" level to the end

```{r}
#| label: bar-chart-freq-lump-other-end
#| echo: true
#| eval: false

diamonds_100 |> 
ggplot(aes(x = 
        fct_lump_n(clarity, n = 4) |> 
        fct_infreq() |> 
        fct_relevel("Other", 
            after = Inf))) +
geom_bar() +
labs(title = "Bar chart with freq. 
    ordering + lumping + relevel",
    x = "Brilliantly arranged factor variable")
```


## Modeling categorical data

**Factor order matters in models!** The first level is the "dummy" or reference category. Output is only interpretable in relation to that first level.

Example: using `mtcars` data, predict `mpg` using `cyl` (number of cylinders) as a factor variable.

Fit regression model with `cyl` as factor

```{r}
#| label: model-reference-level-default

summary(
    lm(mpg ~ factor(cyl), data = mtcars)
)
```

Use `fct_rev()` to change reference level

```{r}
#| label: model-reference-level-reverse

summary(
    lm(mpg ~ fct_rev(factor(cyl)), data = mtcars)
)

```
