
```{r}
#| label: setup
#| include: false

library(tidyverse)
```

# Strings

## Character data

-   [Character data]{.keyword}: R datatype for text
-   [Character]{.keyword}: Smallest unit of text
-   [String]{.keyword}: Sequence of zero or more characters
-   [Character vector]{.keyword}: Collection of one or more strings
    -   A single string is a character vector of length 1

Use single `'` or double `"` quotes to define strings:

```{r}
#| label: create-strings

favorite_class <- "d2m-R"

typeof(favorite_class) # data type
class(favorite_class) # object class
length(favorite_class) # no. elements
str_length(favorite_class) # no. characters
```

Use the combine function [c()]{.fun} to create vectors of strings:

```{r}
#| label: create-char-vector

favorite_class_units <- c("d", "2", "m", "-", "R")

typeof(favorite_class_units) # data type
class(favorite_class_units) # object class
length(favorite_class_units) # no. elements
str_length(favorite_class_units) # no. char.
```

::: notes
`typeof()` vs. `class()`: `typeof()` gives the low-level data type (character), while `class()` gives the high-level object class (character vector). A single string is a character vector of length 1, while multiple strings form a character vector of length > 1.
With strings this will be the same (character), but you'll see the difference with factors.
:::

## Define Examples

*Define strings and vectors to use as examples in later slides:*

Use single `'` or double `"` quotes to define strings:

```{r}
#| label: create-examples

sentence <- 'The QUICK brown fox jumps over the LAZY dog.'
favorite_class <- "d2m-R"
favorite_class_units <- c("d", "2", "m", "-", "R")

fruit <- c("apple", "banana", "cherry", "date", "elderberry")
greeting <- c("hello", "world")
farewell <- c("goodbye", "everyone")
abed_ratings <- c("cool", "cool cool cool", "cool cool", "cool", "sarcastic: ok", "evil: hot")
```



## Base R vs tidyverse

**Base R**: Inconsistent naming

- `grep`, `grepl`, `sub`, `gsub`, `regexpr`, `gregexpr`
- Inconsistent argument order

**`stringr`**: Consistent `str_*` pattern

- `str_detect`, `str_replace`, `str_replace_all`
- Always: string first, pattern second
- Vectorized by default

Like most things in the tidyverse, `stringr` improves upon base r options by providing a consistent naming scheme and argument order.
Contrast base R's inconsistent naming (`grep`, `sub`, `gsub`) with `stringr`'s consistent `str_*` pattern. `stringr` is vectorized by default and uses consistent argument ordering (string first, pattern second).

"Vectorized by default" means that `stringr` functions are designed to work with vectors of strings without needing to use `sapply()` or loops. You can pass a character vector to a `stringr` function, and it will operate on each element of the vector automatically, returning a vector of results. In contrast, many base R string functions are not vectorized and require additional steps to apply them to each element of a character vector.


### `stringr` function families

- **Combine**: Join strings (`str_c`, `str_flatten`)
- **Detect**: Find patterns (`str_detect`, `str_locate`)
- **Extract**: Pull out text (`str_sub`, `str_extract`)
- **Replace**: Modify text (`str_replace`, `str_remove`)
- **Transform**: Change format (`str_to_upper`, `str_trim`)


# Create and Combine Strings



## Combine: Concatenate

-   [Concatenate]{.keyword}: Join multiple strings together into one string
-   Concatenation functions take *strings* as arguments, not vectors of multiple strings

**Base:**

- `paste()`: space separator by default
- `paste0()`: no separator


**`stringr`:**

- `str_c()`: no separator, `NA`s propagate

Combine [c()]{.fun} vs. concatenate [paste()]{.fun} / [str_c()]{.fun}: 

```{r}
#| label: combine-vs-concatenate

greeting <- c("hello", "world")

paste("hello", "world") # concatenate strings into a string
paste(greeting) # vector of strings treated as one argument, not concatenated

str_c("hello", "world") # concatenate strings into a string (no sep.)
str_c(greeting) # vector of strings treated as one argument, not concatenated


```


Compare separator behavior:

```{r}
#| label: compare-separator
paste("hello", "world") # default space separator
str_c("hello", "world", sep = " ") # custom space separator

paste0("hello", "world") # no separator
str_c("hello", "world") # no separator
```

Compare NA handling:

```{r}
#| label: compare-na-propagate

paste("hello", NA) # NA treated as "NA" (string)
str_c("hello", NA) # NA propagates to result
```


Both `paste()` and `str_c()` are vectorized:

```{r}
#| label: show-vectorization

strings1 <- c("hello", "goodbye")
strings2 <- c("world", "everyone")

paste(strings1, strings2) # vectorized with space separator
str_c(strings1, strings2) # vectorized with no separator

```


### NA propagation

In `paste()`, if any argument is `NA`, it is treated as the string "NA". In contrast, `str_c()` propagates `NA` values, meaning that if any argument is `NA`, the result will also be `NA`. This allows you to preserve missing values in your data instead of converting them to a string representation.


## Combine: Collapse

-   [Collapse]{.keyword}: Join elements of a string vector into a single string
-   Collapse-type functions take *vectors* as arguments, not individual strings


**Base:** Base `paste()` concatenates with `sep` arg and collapses vectors with `collapse` arg

Compare `sep` and `collapse` behavior:

```{r}
#| label: sep-collapse

paste(fruit, sep = ", ") # sep does not affect vector
paste(fruit, collapse = ", ") # collapse with comma

# sep ignored when collapse is used)
paste(fruit, sep = "SEPARATOR!!!", collapse = ", ") 

```

**`stringr`:** `str_flatten()` collapses vectors with `collapse` arg, no `sep` arg

Use `str_flatten()` with different (collapsing) separators:

```{r}
#| label: str-flatten-examples

str_flatten(fruit) # default no separator
str_flatten(fruit, collapse = ", ") # collapse with comma
str_flatten(fruit, collapse = " ") # collapse with space
```



### Separators vs. Collapse

Separators are for combining multiple strings, each provided as a separate argument. Collapse is for combining elements of a single vector into one string.

- `sep`: Specifies the string to use to separate individual elements when concatenating multiple strings together. It is used when you want to combine multiple strings into a single string with a specific separator between them. For example, `str_c("hello", "world", sep = " ")` will produce "hello world".
- `collapse`: Specifies the string to use to separate elements when collapsing a vector of strings into a single string. It is used when you have a vector of strings and want to combine them into one string with a specific separator between each element. For example, `str_flatten(c("apple", "banana", "cherry"), collapse = ", ")` will produce "apple, banana, cherry".



## Interpolate

-   [Interpolate]{.keyword}: Embed variable values or expressions directly within strings

**Base:** Interpolate with placeholders using `sprintf()`

-   Use placeholders: `%s` (string), `%d` (integer), `%f` (float)

```{r}
#| label: sprintf-example

name <- "Natalie"
age <- 67

sprintf("My name is %s and I am %d years old.", name, age)
```


**`stringr`:** Interpolate directly with `str_glue()`

-   Evaluate R expressions directly in strings using `{}`

```{r}
#| label: str-glue-example

name <- "Natalie"
age <- 67

str_glue("My name is {name} and I am {age} years old.")
```


-   Include simple calculations:

```{r}
#| label: str-glue-calculation

x <- 10
str_glue("The value of x is {x} and x squared is {x^2}.")

```


-   Include embedded R code:

```{r}
#| label: str-glue-functions

a <- 5
b <- 3

str_glue("The sum of a and b is {a + b} and their mean is {mean(c(a,b))}.")
```




# String Detection and Matching

## Detect: Identify patterns

**Does the pattern exist?** Detect patterns in character vectors of any length (i.e., vectorized or single strings).

**Base:** Find pattern (T/F) with `grepl()` (=grep **l**ogical)

- `grepl()`

```{r}
#| label: detect-grepl

grepl("a", "banana") # returns TRUE
grepl("z", "banana") # returns FALSE
grepl("err", fruit) # returns TRUE for strings containing "err"

```

**`stringr`:** Find pattern (T/F) with `str_detect()`

```{r}
#| label: detect-str-detect

str_detect("banana", "a") # returns TRUE
str_detect("banana", "z") # returns FALSE
str_detect(fruit, "err") # returns TRUE for strings containing "err"
str_detect(fruit, "err", negate = TRUE) # returns TRUE for strings NOT containing "err"
```


Use `str_detect()` in tidy pipelines:

```{r}
#| label: detect-filter

# Using str_detect() in filter()
data.frame(fruit) |> 
  filter(str_detect(fruit, "a")) # keep fruits containing "a"

```


Detection is case-sensitive by default, but can be made case-insensitive with `regex()` wrapper and  `ignore_case` arg:

```{r}
#| label: detect-case-insensitive

str_detect("Banana", "a") # case-sensitive, returns TRUE
str_detect("Banana", "A") # case-sensitive, returns FALSE

# Use regex to make case-insensitive, returns TRUE
str_detect("Banana", 
    regex("A", ignore_case = TRUE)) 

```


Note: `grep` detection can be used in pipelines too, but `str_` functions are more consistent and easier to read as things get more complex.

`grep` doesn't have a simple way of negating the pattern (i.e., find strings that do NOT match) like str_detect, but you can use the `!` operator to invert logical output.



## Detect: Locate patterns

**Where does the pattern appear?**

**Base:** Use `regexpr()` or `gregexpr()` to find position of first match or all matches, respectively

```{r}
#| label: detect-regexpr

# regexpr() returns position of first match only
base_first_match <- regexpr("an", "banana") 
base_first_match
class(base_first_match) # integer with attributes

# gregexpr() returns positions of all matches
base_all_match <- gregexpr("an", "banana") 
base_all_match
class(base_all_match) # list of integer vectors with attributes

```


**`stringr`:** Use `str_locate()` or `str_locate_all()` to find start/end positions of first match or all matches, respectively

```{r}
#| label: detect-str-locate

# str_locate() returns matrix with start/end of first match
stringr_first_match <- str_locate("banana", "an") 
stringr_first_match
class(stringr_first_match) # matrix with start/end columns

# str_locate_all() returns list of matrices with start/end of all matches
stringr_all_match <- str_locate_all("banana", "an")
stringr_all_match
class(stringr_all_match) # list of matrices with start/end columns

```


::: notes
**Pattern location**: Compare `gregexpr()` output (list of positions) with `str_locate()` and `str_locate_all()` (matrix/list of matrices with start/end).
:::

## Detect: Count patterns

**How many times does the pattern appear?**

**Base:** Use `gregexpr()` + `lengths()` to count all matches

```{r}
#| label: detect-count-base

# Use gregexpr() to find all matches, then lengths() to count
lengths(gregexpr("an", "banana"))
```

**`stringr`:** Use `str_count()` to count occurrences of pattern in each string

```{r}
#| label: detect-count-str-count

str_count("banana", "an") # counts occurrences in single string
```


Count words (using regex pattern for word boundaries: `\b` = word boundary, `\w+` = one or more word characters):

```{r}
#| label: count-words

sentence <- "The quick brown fox jumps over the lazy dog."

# Count words using word boundary pattern
str_count(sentence, "\\b\\w+\\b") # counts words by matching word boundaries

```

Vectorized counting across a vector of strings:

```{r}
#| label: count-vector

# Count occurrences of "cool" in each string of abed_ratings vector
str_count(abed_ratings, "cool")
```


# String Manipulation

## Extract: Subset by pattern

**Extract strings that match pattern**

**Base:** Use `grep()` with `value = TRUE` to return matching strings

```{r}
#| label: grep-value-true

grep("a", fruit, value = TRUE) # returns strings containing "a"
```


**`stringr`:** Use `str_subset()` to return strings that match pattern

```{r}
#| label: str-subset

str_subset(fruit, "a") # returns strings containing "a"
```


`str_subset()` is a shortcut for `str_detect()` + subsetting

```{r}
#| label: str-subset-vs-detect

# detect matches returns a logical vector
# then subset with [] indexing on the TRUE values
fruit[str_detect(fruit, "a")]
```


## Extract: Subset by position

**Extract by position**

**Base:** Use `substr()` to extract substring by character positions

`substr(x, start, stop)`

- 1-indexed
- Positive indices only

```{r}
#| label: substr-example

substr(sentence, 1, 9) # extract characters 1-9

```


**`stringr`:** Use `str_sub()` to extract substring by character positions, supports negative indexing

`str_sub(x, start, end)`

- 1-indexed
- **Negative indexing from end**

```{r}
#| label: str-sub-example

str_sub(sentence, 1, 9) # extract characters 1-9
str_sub(sentence, -4, -1) # extract last 4 characters
```


Bonus: interpolate with `str_glue()` to create dynamic substrings:

```{r}
#| label: str-glue-substring

pos_start <- 1
neg_end <- -1

str_glue("Abridged sentence: {str_sub(sentence, pos_start, 9)} {str_sub(sentence, -4, neg_end)}")
```


## Replace patterns

**Replace patterns in strings**

| Base R | `stringr` | Behavior |
| --- | --- | --- |
| `sub()` | `str_replace()` | Replace **first** match |
| `gsub()` | `str_replace_all()` | Replace **all** matches |

**Base:** `sub()` replaces first match, `gsub()` replaces all matches

```{r}
#| label: base-replace

sub("a", "X", "banana") # replaces first "a" with "X"
gsub("a", "Y", "banana") # replaces all "a" with "Y"
gsub("err", "Z", fruit) # replaces "err" with "Z" in all strings containing "err"
```


**`stringr:`** Use `str_replace()` to replace first match, `str_replace_all()` to replace all matches

```{r}
#| label: str-replace

str_replace("banana", "a", "X") # replaces first "a" with "X"
str_replace_all("banana", "a", "Y") # replaces all "a" with "Y"
str_replace_all(fruit, "err", "Z") # replaces "err" with "Z" in all strings containing "err"
```

Use regex patterns in replacement:

```{r}
#| label: regex-replace

# replace all words with "WORD"
str_replace_all(sentence, "\\b\\w+\\b", "WORD") 

# Relace all vowels with "*" and spaces with "_"
str_replace_all(sentence, "[aeiouAEIOU]", "*") |> 
  str_replace_all(" ", "_")

```



## Transform: Trim whitespace

**Clean up messy text**

**Base:** Use `trimws()` to remove leading and trailing whitespace

```{r}
#| label: trimws-example

messy_string <- "   Hello,   World!   "
trimws(messy_string) # removes leading and trailing whitespace
trimws(messy_string, which = "left") # removes leading whitespace only)
trimws(messy_string, which = "right") # removes trailing whitespace only)
```


**`stringr`:** Use `str_trim()` to remove leading and trailing whitespace, `str_squish()` to also collapse internal whitespace

```{r}
#| label: str-trim-squish

str_trim(messy_string) # removes leading and trailing whitespace
str_trim(messy_string, side = "left") # removes leading whitespace only
str_trim(messy_string, side = "right") # removes trailing whitespace only

# removing both leading and trailing
# AND ALSO collapse any internal whitespace to a single space
str_squish(messy_string)

```



::: notes
**Trimming whitespace**: Compare `trimws()` with `str_trim()` (removes leading/trailing) and `str_squish()` (also collapses internal whitespace).
::: 


## Transform: Convert case

| Base R | `stringr` | Result |
| --- | --- | --- |
| `toupper()` | `str_to_upper()` | ALL CAPS |
| `tolower()` | `str_to_lower()` | all lowercase |
| — | `str_to_title()` | Title Case Each Word |
| — | `str_to_sentence()` | Sentence case |

**Base:** Use `toupper()` and `tolower()` for case conversion

```{r}
#| label: base-case-conversion

toupper(sentence) # converts to ALL CAPS
tolower(sentence) # converts to all lowercase
```


**`stringr:`** Use `str_to_upper()`, `str_to_lower()`, `str_to_title()`, and `str_to_sentence()` for case conversion

```{r}
#| label: str-case-conversion

str_to_upper(sentence) # converts to ALL CAPS
str_to_lower(sentence) # converts to all lowercase
str_to_title(sentence) # converts to Title Case Each Word
str_to_sentence(sentence) # converts to Sentence case
```



### Locale-aware

Some `stringr` functions have local-aware case conversion like Turkish dotted/dotless i: `str_to_upper("i", locale = "tr")` returns "İ" (dotted capital I) and `str_to_lower("I", "tr")` returns "ı" (dotless lowercase i) in Turkish locale, while base R's `toupper("i")` and `tolower("I")` return "I" and "i" respectively regardless of locale.


# Regular Expressions

## Regex overview

**Literal characters**: Match exactly

**Metacharacters** have special meaning, like defining what kinds of characters to recognize, how many times to match, or where in the string to look. Common metacharacters:

| Metacharacter | Meaning |
| --- | --- |
| `.` | Any single character except newline |
| `*` | 0 or more of the preceding element |
| `+` | 1 or more of the preceding element |
| `?` | 0 or 1 of the preceding element |
| `[]` | Character class (matches any one character inside) |
| `{}` | Quantifier (specifies number of occurrences) |
| `^` | Start of string |
| `$` | End of string |
| `\` | Escape character (treat next character literally) |
| `|` | Alternation (aka "or" matches either the pattern before or after) |

**Escaping** allows you to match metacharacters as literal characters by preceding them with the escape metacharacter.

- Match literal period: `\.` → `"\\."` in R
- Match literal asterisk: `\*` → `"\\*"` in R

```{r}
#| label: regex-literal-vs-metacharacter

# Literal match with metacharacter
str_detect("file.txt", "\\.") # matches literal period
str_detect("filetxt", "\\.") # no match, no period

# Metacharacter match
str_detect("filetxt", ".") # matches any character
str_detect("", ".") # no match, empty string has no characters

```

## Character classes & quantifiers

**Character classes** match any one character from a set. Common character classes:

| Class | Meaning |
| --- | --- |
| `[abc]` | a, b, or c |
| `[^abc]` | NOT a, b, or c |
| `[a-z]` | any lowercase letter |
| `\d` | digit |
| `\w` | word character (letter, digit, or underscore) |
| `\s` | whitespace character (space, tab, newline) |

```{r}
#| label: character-classes

str_detect("apple", "[aeiou]") # matches if contains any vowel
str_detect("apple", "[^aeiou]") # matches if contains any non-vowel
str_detect("apple", "\\d") # matches if contains any digit
str_detect("apple", "\\w") # matches if contains any word character (letter, digit, or underscore)
str_detect("apple", "\\s") # matches if contains any whitespace
```


**Quantifiers** specify how many times the preceding element should be matched:

| Quantifier | Meaning |
| --- | --- |
| `*` | 0 or more of the preceding element |
| `+` | 1 or more of the preceding element |
| `?` | 0 or 1 of the preceding element |
| `{n}` | exactly n of the preceding element |
| `{n,}` | n or more of the preceding element |
| `{n,m}` | between n and m of the preceding element |

```{r}
#| label: quantifiers

str_detect("aaab", "a*") # matches 0 or more a's (matches entire string)
str_detect("aaab", "a+") # matches 1 or more a's (matches "aaab")
str_detect("aaab", "a?") # matches 0 or 1 a (matches "a" at the start)
str_detect("aaab", "a{3}") # matches exactly 3 a's (matches "aaa")
str_detect("aaab", "a{2,}") # matches 2 or more a's (matches "aaab")
str_detect("aaab", "a{2,3}") # matches between 2 and 3 a's (matches "aaab" because it has 3 a's, but would not match "aab" because it only has 2 a's)

```

Use quantifiers in replacement functions to specify how many occurrences to replace:

```{r}
#| label: other-str-functions

str_replace_all("aaab", "a{2,3}", "X") # replaces 2 or 3 a's with "X" (result: "Xb")
str_locate_all("aaab", "a{2,3}") # locates positions of 2 or 3 a's (result: list of matrices with start/end positions)
str_count("aaab", "a{2,3}") # counts occurrences of 2 or 3 a's (result: 1)

```


### Greedy vs lazy matching (`.*?` vs `.*`)

Greedy matches as much as possible, while lazy matches as little as possible. In the example below, the greedy pattern `.*` captures everything between "Start" and "end", while the lazy pattern `.*?` captures only the text in between without consuming everything.

```{r}
#| label: greedy-vs-lazy

double_dog <- "The quick brown fox jumps over the lazy dog. The dog was displeased."

# Greedy match (matches everything between "The" and "dog")
str_extract(double_dog, "The.*dog") # Result: "The quick brown fox jumps over the lazy dog. The dog"

# Lazy match (matches the shortest string between "The" and "dog")
str_extract(double_dog, "The.*?dog") # Result: "The quick brown fox jumps over the lazy dog"

```


## Anchors & boundaries

Use **anchors** to specify where in the string to look for a match:

| Anchor | Meaning |
| --- | --- |
| `^` | Start of string |
| `$` | End of string |
| `\b` | Word boundary (position between a word character and a non-word character
| `\B` | Non-word boundary (position between two word characters or two non-word characters) |

Note that the carat `^` has two uses: as an anchor for the start of the string, and as a negation symbol when used inside square brackets for character classes (e.g., `[^abc]` means "not a, b, or c"). The meaning depends on the context in which it is used.

```{r}
#| label: anchor-start

str_detect("banana", "^ba") # matches "ba" at start
str_detect("banana", "^na") # does not match "na" at start
```


Match entire string with `^` and `$`:

```{r}
#| label: anchor-start-end

str_detect("banana", "^banana$") # matches entire string "banana"
str_detect("banana", "^ban$") # does not match because "ban" is not the entire string
```


Use word boundaries for whole word matching:

```{r}
#| label: word-boundary

sea_cat <- "The cat is on the catamaran. She's a weird cat."

str_count(sea_cat, " cat ") # matches "cat" surrounded by spaces
str_count(sea_cat, "\\bcat\\b") # matches "cat" as a whole word
str_count(sea_cat, "cat") # matches "cat" as part of "cat" and "catamaran"
```


Validation pattern example (e.g., email, phone)

```{r}
#| label: validation-pattern

# Simple email validation pattern: one or more word characters, @, one or more word characters, ., two or more letters
email_pattern <- "^\\w+@\\w+\\.\\w{2,}$"

str_detect("name@school.edu", email_pattern) # returns TRUE for valid email format
str_detect("invalid email@spammer,com", email_pattern) # returns FALSE for invalid email format

```



## Capture groups & back-references

**Capture** groups to extract specific parts of a matched pattern and **back-reference** them in replacement or later in the regex. 

Use parentheses `()` to define capture groups, and backreferences like `\1`, `\2`, etc. to refer to those groups in the replacement string.

- `()`: Create capture group
- `\1`, `\2`, etc.: Reference captured groups in replacement

Use `back-references` to reorder captured groups:

```{r}
#| label: backreferences

# Capture the two animals (fox, dog) and swap their placement in the string
str_replace(sentence, "(fox)(.*)(dog)", "\\3\\2\\1")

```

Example: Reorder US dates `MM/DD/YYYY` to `YYYY-MM-DD` using capture groups and back-references:

```{r}
#| label: date-reorder

new_year <- "The date is 12/31/2025. Happy New Year!"
birthday <- "The date is 6/7/25. Happy birthday!"

# Capture month, day, year and reorder to YYYY-MM-DD
str_replace(new_year, "(\\d{2})/(\\d{2})/(\\d{4})", "\\3-\\1-\\2") # matches because month, day, and year are expected number of digits
str_replace(birthday, "(\\d{1,2})/(\\d{1,2})/(\\d{2,4})", "\\3-\\1-\\2") # account for the possibility of 1-digit month/day and 2-digit year


```


## Regex Challenge

Create a function for the date reordering that can handle both 1 or 2 digit month/day and 2 or 4 digit year formats as input but always deliver a consistent `YYYY-MM-DD` output format. Use capture groups to extract the components and back-references to reorder them in the desired format. Test your function on various date strings to ensure it works correctly.

```{r}
#| label: regex-challenge

# Create function `reorder_date()`
reorder_date <- function(date_string) {
    
  # Define regex pattern to capture month, day, and year with flexible digit counts

  
  # Use str_replace to reorder the date components to YYYY-MM-DD

}

# Test with multiple string types
reorder_date("4/5/2023") # Expected: "2023-04-05"



## Even challengier: Handle multiple date formats by adding arguments for the user
## to specify the input and/or format 


```
 
 